Combined effects of protein expression variance and correlation on multicomponent systems

### 2020-03-03 Analysis.Rmd
update

for(i in seq_along(files)){
  if(i == 1 & exists('pepRaw')){
    rm(pepRaw)
  } else if(!exists('pepRaw')){
    pepRaw = read_tsv(file = files[i]) %>% mutate(file = files[i], massID = 1:nrow(.))
  } else{
    temp = read_tsv(file = files[i]) %>% mutate(file = files[i], massID = 1:nrow(.))
    colnames(temp) = colnames(pepRaw)
    pepRaw = rbind(pepRaw, temp)
    rm(temp)
  }
}
pepRaw <- pepRaw %>% dplyr::select(file, massID, everything()) %>% filter(!is.na(`126`), 
                                                                  `Quan Usage` == 'Use',
                                                                  Confidence != 'Low')

@@

```{r echo=TRUE, fig.height=10, fig.width=10,message=FALSE, warning=FALSE}
meanSD = pepTidy %>% filter(condition != 'ctrl') %>% 
  group_by(peptide) %>% 
  summarise(mean = mean(intensity, na.rm = T), 
            sd = sd(intensity, na.rm = T) / mean)

# ggplot(meanSD, aes(x = log10(mean), y = sd)) + geom_point() + geom_smooth(method = lm, formula = y ~ poly(x, 2))
```

@@

```{r echo=TRUE, fig.height=5, fig.width=8, message=FALSE, warning=FALSE}
pepTidy = pepTidy %>% group_by(cell) %>% mutate(intensity = intensity / median(intensity, na.rm = T))

@@

```{r echo=TRUE, fig.height=5, fig.width=8, message=FALSE, warning=FALSE}
pepTidy = pepTidy %>% group_by(run, peptide) %>% 
  mutate(intensity = intensity / median(intensity, na.rm = T)) %>% ungroup()

@@

There are

@@

```{r echo=TRUE, fig.height=5, fig.width=8, message=FALSE, warning=FALSE}
pepTidy %>% dplyr::filter(abs(log2(intensity)) < 2*sd(log2(pepTidy$intensity), na.rm = T))

ggplot(pepTidy, aes(log2(intensity), colour = cell)) + geom_density() + theme_bw() + xlim(-4,4)
```

@@

@@

mypal = rev(colorRampPalette(brewer.pal(11, 'RdBu'))(201))
k = 2
map = pheatmap(cor, show_colnames = F, show_rownames = F, color = mypal, breaks = seq(-1,1, by = 0.01), cutree_cols = k, cutree_rows = k, legend = F)
#map

tibble(Uniprot = protImpute$Uniprot, Cluster = cutree(map$tree_row, k = k)) %>% write_csv('/Users/kylekovary/Downloads/Map.csv')

@@

pcaDF = tibble(cell = rownames(pca$rotation), PC1 = pca$rotation[,1], PC2 = pca$rotation[,2])
pcaDF = pcaDF %>% separate(cell, into = c('run','label','condition'), remove = F) %>% print()
ggplot(pcaDF, aes(x = PC1, y = PC2, colour = condition)) + geom_point(size = 5)

@@

```{r eval=F}
columns(org.Xl.eg.db)
#mapIds(org.Xl.eg.db, keys = 'A0JPG5', keytype = 'UNIPROT', column = 'GOALL', multiVals = 'list')

@@

allGO <- allGO %>% group_by(GOterms) %>% 
  dplyr::mutate(cors = corVector(Uniprot, cor) %>% as_tibble() %>% as.list(),
                cv = selectCV(Uniprot)) %>% ungroup() %>%
  rowwise() %>% mutate(medCors = mean(cors, na.rm = TRUE),
                       medCV = mean(cv, na.rm = TRUE)) %>%
  ungroup() %>%
filter(length(Uniprot) > 3,
       !is.na(medCors),
       !is.na(medCV))

@@

protVar <- protImpute %>% gather(key = 'cell', value = 'abundance', 2:ncol(protImpute)) %>%
  separate(cell, into = c('run','label','condition'), remove = F) %>% 
  filter(condition != 'ctrl') %>% group_by(Uniprot) %>%
  summarise(cv = sd(abundance)/mean(abundance))

ggplot(protVar, aes(x = cv)) + geom_histogram(bins = 50) + theme_bw()

@@

```{r message=FALSE, warning=FALSE}
allGO[grep('ribosom',allGO$GO),]
glyc <- filter(allGO, GOterms == 'GO:0005840')$Uniprot %>% unlist()

@@

## Cor vs CV
```{r}
test = allGO %>% filter(length(Uniprot) > 3, !is.na(medCors), !is.na(medCV))

@@

allGO1 <- allGO %>% filter(!is.na(GOterms)) %>% rowwise() %>% 
  mutate(n = length(unlist(Uniprot)),
         totVar = totVarGroup(Uniprot),
         normVar = totVar / n) %>%
  filter(n > 3)

@@

ggplot(allGO1, aes(x = normVar)) + geom_histogram(bins = 30) + 
  geom_vline(xintercept = c(0.15, 0.1), colour = 'red', linetype = 'dashed') +
  theme_bw()

@@

### Is lower variance compensating for high correlations? How much of a role is negative correlation playing in reducing the total variance?
```{r}
dat <- left_join(ungroup(test), allGO1, by = "GOterms") %>%
  dplyr::select(GOterms, GO.x, ontology.x, Uniprot.x, cors.x, cv.x, n.x,medCors, medCV, normVar)

dat$score <- 'Med'
dat[dat$normVar > 0.145, 'score'] <- 'High'
dat[dat$normVar < 0.115, 'score'] <- 'Low'

@@

ggplot(filter(dat, n.x > 5), aes(x = medCors, y = medCV, colour = score)) + 
  geom_point(alpha = 0.75) + scale_color_manual(values=c("blue", "red", "grey"))
# + #scale_colour_gradient(low="red", high="blue") +
#  geom_text_repel(data = filter(dat, score != 'Med'), aes(label = GOterms), colour = 'black', size = 1)

ggplot(filter(dat, n.x > 5, score != 'Med', ontology.x != 'CC'), aes(x = medCors, y = medCV, colour = normVar)) + 
  geom_point() + scale_colour_gradient(low="red", high="blue") + 
  geom_smooth(method = 'lm') + facet_wrap(~score, scales = 'free') +
  geom_text_repel(data = filter(dat, n.x > 5, score != 'Med'), aes(label = GO.x), colour = 'black', size = 2)

plot_ly(filter(dat, n.x > 5), z = ~normVar, y = ~medCV, x = ~medCors, color = ~score, colors = c("red", "blue", "grey")) %>%
  add_markers() %>%
  layout(scene = list(zaxis = list(title = 'Normalized Variance'),
                     yaxis = list(title = 'Median CV'),
                     xaxis = list(title = 'Median Correlation')))

ggplot(filter(dat, n.x > 5), aes(y = normVar, x = medCV)) + geom_point(alpha = 0.75)
ggplot(filter(dat, n.x > 5), aes(y = normVar, x = medCors)) + geom_point(alpha = 0.75)

library(GGally)
p <- ggpairs(filter(dat, n.x > 5), columns = c('medCV', 'medCors','normVar'), 
        ggplot2::aes(colour = score, alpha = 0.5))

>>>>

for(i in seq_along(files)){
  if(i == 1 & exists('pepRaw')){
    rm(pepRaw)
  } 
  if(!exists('pepRaw')){
    pepRaw = read_tsv(file = files[i]) %>% mutate(file = files[i], massID = 1:nrow(.))
  } else{
    temp = read_tsv(file = files[i]) %>% mutate(file = files[i], massID = 1:nrow(.))
    colnames(temp) = colnames(pepRaw)
    pepRaw = rbind(pepRaw, temp)
    rm(temp)
  }
}
pepRaw <- pepRaw %>% dplyr::select(file, massID, everything()) %>% filter(!is.na(`126`), 
                                                                  `Quan Usage` == 'Use',
                                                                  Confidence != 'Low')

@@

@@

```{r echo=TRUE, fig.height=5, fig.width=8, message=FALSE, warning=FALSE}
pepTidy = pepTidy %>% group_by(cell) %>% mutate(intensity = intensity / mean(intensity, na.rm = T))

@@

```{r echo=TRUE, fig.height=5, fig.width=8, message=FALSE, warning=FALSE}
pepTidy = pepTidy %>% group_by(run, peptide) %>% 
  mutate(intensity = intensity / mean(intensity, na.rm = T)) %>% ungroup()

@@

@@

```{r echo=TRUE, fig.height=5, fig.width=8, message=FALSE, warning=FALSE}
pepTidy <- pepTidy %>% dplyr::filter(abs(log2(intensity)) < 2*sd(log2(pepTidy$intensity), na.rm = T))

#ggplot(pepTidy, aes(log2(intensity), colour = cell)) + geom_density() + theme_bw() + xlim(-4,4)
```

@@

protVar <- protImpute %>% gather(key = 'cell', value = 'abundance', 2:ncol(protImpute)) %>%
  separate(cell, into = c('run','label','condition'), remove = F) %>% 
  filter(condition != 'ctrl') %>% group_by(Uniprot) %>%
  summarise(var = var(abundance),
            sd = sd(abundance),
            mean = mean(abundance),
            cv = sd / mean)

ggplot(protVar, aes(x = sd)) + geom_histogram(bins = 50) + theme_bw()

@@

mypal = rev(colorRampPalette(brewer.pal(11, 'RdBu'))(201))
k = 1
map = pheatmap(cor, show_colnames = F, show_rownames = F, color = mypal, breaks = seq(-1,1, by = 0.01), cutree_cols = k, cutree_rows = k, legend = F)
map

cov = cov(t(log2(protImpute[c(1:ncol(protImpute))[-c(1,grep('ctrl',colnames(protImpute)))]])), use = 'pairwise.complete.obs')
colnames(cov) = protImpute$Uniprot
rownames(cov) = protImpute$Uniprot

#tibble(Uniprot = protImpute$Uniprot, Cluster = cutree(map$tree_row, k = k)) %>% write_csv('/Users/kylekovary/Downloads/Map.csv')

@@

pcaDF = tibble(cell = rownames(pca$rotation), PC1 = pca$rotation[,1], PC2 = pca$rotation[,2])
pcaDF = pcaDF %>% separate(cell, into = c('run','label','condition'), remove = F) %>% print()
ggplot(pcaDF, aes(x = PC1, y = PC2, colour = condition)) + stat_ellipse() + 
  geom_point(size = 5)

@@

```{r eval=TRUE}
#columns(org.Xl.eg.db)
#mapIds(org.Xl.eg.db, keys = 'A0JPG5', keytype = 'UNIPROT', column = 'GOALL', multiVals = 'list')

@@

allGO <- allGO %>% group_by(GOterms) %>% 
  dplyr::mutate(cors = corVector(Uniprot, cor) %>% as_tibble() %>% as.list(),
                covs = corVector(Uniprot, cov) %>% as_tibble() %>% as.list(),
                cv = selectCV(Uniprot)) %>% ungroup() %>%
  rowwise() %>% mutate(medCors = mean(cors, na.rm = TRUE),
                       medCV = mean(cv, na.rm = TRUE)) %>%
  ungroup() %>%
filter(n > 3,
       !is.na(medCors),
       !is.na(medCV))

allGO <- allGO %>% rowwise %>% mutate(totVar = sqrt(sum(cv) + sum(covs)),
                             normVar = totVar / n)

@@

@@

```{r message=FALSE, warning=FALSE}
allGO[grep('glycoly',allGO$GO),]
glyc <- filter(allGO, GOterms == 'GO:0006096')$Uniprot %>% unlist()

@@

```{r}
x <- tibble(cor = c(corVect$cor,glycMat[lower.tri(glycMat)]),
       label = c(rep('All', nrow(corVect)),
                 rep('Gly',length(glycMat[lower.tri(glycMat)]))))

ggplot(x, aes(x = cor, fill = label, color = label)) + geom_density(alpha = 0.5) + geom_rug(data = filter(x, label == 'Gly'))

## Cor vs CV
```{r}
test <- allGO %>% filter(n > 3, !is.na(medCors), !is.na(medCV))

@@

allGO1 <- allGO %>% filter(!is.na(GOterms)) %>% rowwise() %>%
  mutate(totVar = totVarGroup(Uniprot),
         normVar = totVar / n) %>%
  filter(n > 3) %>% ungroup()

@@

ggplot(allGO, aes(x = normVar)) + geom_histogram(bins = 30) + 
  geom_vline(xintercept = c(0.15, 0.1), colour = 'red', linetype = 'dashed') +
  theme_bw()

@@

### Is lower variance compensating for high correlations? How much of a role is negative correlation playing in reducing the total variance?
```{r}
dat <- left_join(ungroup(test), allGO1, by = "GOterms") %>%
  dplyr::select(GOterms, GO.x, ontology.x, Uniprot.x, cors.x, cv.x, n.x,medCors.x, medCV.x, normVar) %>%
  dplyr::rename(GO = GO.x, ontology = ontology.x, Uniprot = Uniprot.x, cors = cors.x, cv = cv.x,
                n = n.x, medCors = medCors.x, medCV = medCV.x) %>%
  mutate(score = ifelse(normVar > 0.145,
                              'High',
                              ifelse(normVar < 0.115,
                                     'Low',
                                     'Med')))

# dat$score <- 'Med'
# dat[dat$normVar > 0.145, 'score'] <- 'High'
# dat[dat$normVar < 0.115, 'score'] <- 'Low'

@@

ggplot(filter(dat, n > 5), aes(x = medCors, y = medCV, colour = score)) + 
  geom_point(alpha = 0.75) + scale_color_manual(values=c("blue", "red", "grey"))
# + #scale_colour_gradient(low="red", high="blue") +
#  geom_text_repel(data = filter(dat, score != 'Med'), aes(label = GOterms), colour = 'black', size = 1)

ggplot(filter(dat, n > 5, score != 'Med', ontology != 'CC'), aes(x = medCors, y = medCV, colour = normVar)) + 
  geom_point() + scale_colour_gradient(low="red", high="blue") + 
  geom_smooth(method = 'lm') + facet_wrap(~score, scales = 'free') +
  geom_text_repel(data = filter(dat, n > 5, score != 'Med'), aes(label = GO), colour = 'black', size = 2)

plot_ly(filter(dat, n > 5), z = ~normVar, y = ~medCV, x = ~medCors, color = ~score, colors = c("red", "blue", "grey")) %>%
  add_markers() %>%
  layout(scene = list(zaxis = list(title = 'Normalized Variance'),
                     yaxis = list(title = 'Median CV'),
                     xaxis = list(title = 'Median Correlation')))

ggplot(filter(dat, n > 5), aes(y = normVar, x = medCV)) + geom_point(alpha = 0.75)
ggplot(filter(dat, n > 5), aes(y = normVar, x = medCors)) + geom_point(alpha = 0.75)

library(GGally)
p <- ggpairs(filter(dat, n > 5), columns = c('medCV', 'medCors','normVar'), 
        ggplot2::aes(colour = score, alpha = 0.5))

### 2020-03-03 Manuscript/Manuscript.Rmd
updated
update

  word_document: null

documentclass: article
capsize: normalsize
geometry: margin=1in
fontsize: 11pt
spacing: doublespacing
bibliography: 
  - references.bib
  - knitcitations.bib
csl: cell.csl
abstract: Protein expression variation leads to phenotypic variance between cells. This has been demonstrated in cell signaling and differentiation decisions. Additionally coordinated expression of proteins between cells can tune signaling pathways either towards a more binary or analog modality. Though there is some evidence in bulk cell measurements and in bacteria that certain heteromeric subunits or metabolic pathways may be expressed in a coordinated fashion (i.e. operons in bacteria), there has not been a direct measurement of coordinated expression of proteins (independent of TFs) in metazoans (vertebrates?). Here, we measure cell-to-cell variability of relative protein abundance using quantitative proteomics of individual *Xenopus laevis* eggs and show that proteins involved in metabolic pathways or members of heteromeric complexes tend to have high correlations with other members of those pathways/complexes. Our previous work highlighted the fact that correlated expression increases the total variation of a pathway, so one would reason that certain pathways or complexes would need to compensate of this extra source of by reducing the variation of expression of these pathways or complexes. To test this we computed total variance score that took into account both the coefficient of variance and correlations between proteins in a pathway and found that the lower 10% of GO terms were highly enriched for metabolic pathways. When we looked at the relationship between CV and R between these GO terms we found a negative relationship between them, demonstrating that increased correlation needs to come at the expense of decreased variance. Simple molecular models of heteromeric complexes and metabolic pathways demonstrate that this tradeoff can result in higher efficiencies in both function and reduced energy waste. Together, our study argues for a control principle whereby coordinated expression of proteins in a pathway can require lower variance in order to reduce the overall pathway variance, enabling accurate control of active complexes or metabolic pathway activity.
---

@@

Variance Sum Law — Independent Case

$$\sigma_{_{x+y}}^2 = \sigma_{_{x}}^2 + \sigma_{_{y}}^2$$

@@

$$\sigma_{_{x+y}}^2 = \sigma_{_{x}}^2 + \sigma_{_{y}}^2 + 2Cov(x,y)$$
$$\rho = \dfrac{Cov(x,y)}{\sigma_{_{x}} \sigma_{_{y}}}$$
$$\sigma_{_{x+y}}^2 = \sigma_{_{x}}^2 + \sigma_{_{y}}^2 + 2\rho \sigma_{_{x}} \sigma_{_{y}}$$

$$\sigma_{{}_{total}} = \sqrt{\sum\limits_{i=1}^{n} \; \sigma_{i}^{2} \; \sum\limits_{j\neq i} \; 2\rho_{_{i,j}}\sigma_{_{i}}\sigma_{_{j}}}$$

$$\varsigma_{{}_{norm}} = \dfrac{\sigma_{{}_{total}}}{n}$$

>>>>

  word_document: null
z
documentclass: article
capsize: normalsize
geometry: margin=1in
fontsize: 11pt
spacing: doublespacing
bibliography: 
  - references.bib
  - knitcitations.bib
csl: cell.csl
abstract: Protein expression variation leads to phenotypic variance between cells. This has been demonstrated in cell signaling and differentiation decisions. Additionally coordinated expression of proteins between cells can tune signaling pathways either towards a more binary or analog modality. Though there is some evidence in bulk cell measurements and in bacteria that certain heteromeric subunits or metabolic pathways may be expressed in a coordinated fashion (i.e. operons in bacteria), there has not been a direct measurement of coordinated expression of proteins (independent of TFs) in metazoans (vertebrates?). Here, we measure cell-to-cell variability of relative protein abundance using quantitative proteomics of individual *Xenopus laevis* eggs and show that proteins involved in metabolic pathways or members of heteromeric complexes tend to have high correlations with other members of those pathways/complexes. Our previous work highlighted the fact that correlated expression increases the total variation of a pathway, so one would reason that certain pathways or complexes would need to compensate of this extra source of by reducing the variation of expression of these pathways or complexes. To test this we computed total variance score that took into account both the coefficient of variance and correlations between proteins in a pathway and found that the lower 10% of GO terms were highly enriched for metabolic pathways. When we looked at the relationship between CV and R between these GO terms we found a negative relationship between them, demonstrating that increased correlation needs to come at the expense of decreased variance. Simple molecular models of heteromeric complexes and metabolic pathways demonstrate that this tradeoff can result in higher efficiencies in both function and reduced energy waste. Together, our study argues for a control principle whereby coordinated expression of proteins in a pathway can require lower variance in order to reduce the overall pathway variance, enabling accurate control of active complexes or metabolic pathway activity.
---

@@

@@

The sum of variance for a group of proteins $X_{1},\dotsc,X_{n}$ with dependent relationships:

$$var\left(\sum_{i=1}^n X_{i}\right) = \sum_{i=1}^{n} var(X_{i}) + 2\sum_{i,j:i<j}^{n} cov(X_{i},X_{j})$$

$$\sigma_{{}_{total}} = \sqrt{var\left(\sum_{i=1}^n X_{i}\right)}$$

$$\sigma_{{}_{norm}} = \dfrac{\sigma_{{}_{total}}}{n}$$
